---
title: "SQL exploration"
author: "Mark Gallivan"
output: 
  md_document:
    variant: markdown_github
---

SQL exploration of three databases on providers, patients, and healthcare plans. 

### Set-up
```{r setup}

## Load packages 
library(RSQLite) # used create connection to db and get sqlite query instead of bash call
library(knitr) # used to make nice tables 
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)

#tinytex::install_tinytex(force = TRUE) #for latex
## set up connection to  db
SqliteDriver <- dbDriver("SQLite")
DB <- dbConnect(SqliteDriver,
             dbname = "/Volumes/GoogleDrive/My Drive/Admin/data/sqlite.db")
## Through terminal:  
#system("cd /Volumes/GoogleDrive/My Drive/Admin/db/")
#system("sqlite3 $sqlite.db)               
```

## Part I. 

### Question 1. How many rows are in the provider database?

There are a total of `r dbGetQuery(DB, "SELECT COUNT(*) FROM providers")` entries in the provider database.
```{r Q1}
kable(dbGetQuery(DB, "SELECT COUNT(*) as TotalEntries FROM providers"))
```

### Question 2. How many specialists are in the provier database?

There are a total of `r dbGetQuery(DB, "SELECT Count(*) FROM providers WHERE is_pcp = 0")` specialists (or non-primary care physcians) in the database. 
```{r Q2}
kable(dbGetQuery(DB, "SELECT Count(id) as CountNonICPs 
                             FROM providers 
                             WHERE is_pcp = 0"))
```

### Question 3. What states are the providers in?

According to the provider directory, there are providers in five states: Massachusetts, Idaho, Florida, Alaska, and Arkansas. 
```{r Q3}
kable(dbGetQuery(DB, "SELECT DISTINCT 
                             substr(city, INSTR(city, ',')+1, length(city)) as ProviderStates 
                             FROM providers"))
```


#### Question 3a. How many physicians are in each state?

There are 14 physicians in Alaska, 19 in Arkansas, 29 in Florida, 24 in Idaho, and 14 in Massachusetts. 

```{r Q3a}
kable(dbGetQuery(DB, "SELECT substr(city, INSTR(city, ',')+1, length(city)) as States, COUNT(id) as Count
                             FROM providers 
                             GROUP BY States"))
```


#### Question 4. How many rows are in the patient database? 

There are `r dbGetQuery(DB, "SELECT COUNT(*) FROM members")` total entries in the patient database. 
```{r Q4}
kable(dbGetQuery(DB, "SELECT COUNT(*) as CountMembers FROM members"))
```

### Question 5. What states do members live in?

Members live in Arkansas, Alaska, Florida, Idaho, and Massachusetts. 
```{r Q5}
kable(dbGetQuery(DB, "SELECT DISTINCT 
                             substr(city, INSTR(city, ',')+1, length(city)) as MemberStates 
                             FROM members"))
```

#### Question 5a. How many patients are in each state?

As seen in the table below, there are 2036 patients in Alaska, 1974 in Arkansas, 1967 in Florida, 2007 in Idaho, and 2016 in Massachusetts. 
```{r Q5a}
kable(dbGetQuery(DB, "SELECT substr(city, INSTR(city, ',')+1, length(city)) as States, 
                             COUNT(DISTINCT id) as Count 
                             FROM members 
                             GROUP BY States"))
```

#### Question 6. How many members have a primary care provider in each month?

The number of members that have a PCP each month can be found in the table below. However, it is important to note there are `r dbGetQuery(DB, "SELECT COUNT(*) FROM Member_PCP_Spans WHERE provider_id IS NULL")` rows with a missing provider_id, however we will assume this is a data entry error since there is a record in the Member_PCP_Spans table. 

```{r Q6}
kable(dbGetQuery(DB, "WITH NewTable AS 
  (SELECT member_id, CAST(MIN(substr(start_date, 3, 2)) AS INT) 
  as MinStart_Month, CAST(MAX(substr(end_date, 3,2)) AS INT) as MaxEnd_Month 
  FROM Member_PCP_Spans GROUP BY member_id) 
  SELECT SUM(case when MinStart_Month = 1 then 1 else 0 end) as Jan, 
  SUM(case when MinStart_Month <= 2 AND MaxEnd_Month >= 2 then 1 else 0 end) as Feb, 
  SUM(case when MinStart_Month <= 3 AND MaxEnd_Month >= 3 then 1 else 0 end) as Mar, 
  SUM(case when MinStart_Month <= 4 AND MaxEnd_Month >= 4 then 1 else 0 end) as Apr, 
  SUM(case when MinStart_Month <= 5 AND MaxEnd_Month >= 5 then 1 else 0 end) as May, 
  SUM(case when MinStart_Month <= 6 AND MaxEnd_Month >= 6 then 1 else 0 end) as June, 
  SUM(case when MinStart_Month <= 7 AND MaxEnd_Month >= 7 then 1 else 0 end) as Jul, 
  SUM(case when MinStart_Month <= 8 AND MaxEnd_Month >= 8 then 1 else 0 end) as Aug, 
  SUM(case when MinStart_Month <= 9 AND MaxEnd_Month >= 9 then 1 else 0 end) as Sept, 
  SUM(case when MinStart_Month <= 10 AND MaxEnd_Month >= 10 then 1 else 0 end) as Oct, 
  SUM(case when MinStart_Month <= 11 AND MaxEnd_Month >= 11 then 1 else 0 end) as Nov, 
  SUM(case when MinStart_Month <= 12 AND MaxEnd_Month >= 12 then 1 else 0 end) as Dec 
  FROM NewTable")) 
```

#### Question 7. How many members changed their primary care provider just once during the last 6 months of 2018?

There were `r dbGetQuery(DB, "WITH NewTable AS (SELECT member_id, start_date, COUNT(*) as Count_Num FROM Member_PCP_Spans WHERE (substr(start_date,3,2) IN ('07','08','09','10','11','12')) GROUP BY member_id HAVING Count_Num = 1) SELECT COUNT(*) as Count FROM NewTable")` members who changed their PCP exactly once during the last months of 2018. 

```{r Q7}
kable(dbGetQuery(DB, "WITH NewTable AS (SELECT member_id, start_date, COUNT(*) as Count_Num 
           FROM Member_PCP_Spans WHERE (substr(start_date,3,2) IN ('07','08','09','10','11','12')) 
           GROUP BY member_id HAVING Count_Num = 1) SELECT COUNT(*) as Count FROM NewTable"))
```

#### Question 8. In November, how many members are assigned to a Florida provider?

There were `r dbGetQuery(DB, "WITH NewTable AS (SELECT provider_id, member_id, CAST(substr(start_date, 3, 2) AS INT) as Start_Month, CAST(substr(end_date, 3,2) AS INT) as End_Month FROM Member_PCP_Spans) SELECT COUNT(*) FROM NewTable INNER JOIN Providers ON Providers.id = NewTable.provider_id WHERE NewTable.Start_Month <= 11 AND NewTable.End_Month >= 11 AND substr(Providers.city, INSTR(Providers.city, ',')+2, length(Providers.city)) = 'FL'")` members assigned to a provider in Florida in November. However, it's important to note this number only includes PCPs. 

```{r Q8}
kable(dbGetQuery(DB, "WITH NewTable AS (SELECT provider_id, member_id, 
                 CAST(substr(start_date, 3, 2) AS INT) as Start_Month, 
                 CAST(substr(end_date, 3,2) AS INT) as End_Month FROM Member_PCP_Spans) 
                 SELECT COUNT(*) as COUNT FROM NewTable 
                 INNER JOIN Providers ON Providers.id = NewTable.provider_id 
                 WHERE NewTable.Start_Month <= 11 AND NewTable.End_Month >= 11 
                 AND substr(Providers.city, INSTR(Providers.city, ',')+2, length(Providers.city)) = 'FL'"))
```

#### Question 9. How many patients are currently seeing a provider practicing outside of the patient’s city?

There are currently (as of 12/31/18), `r dbGetQuery(DB, "SELECT COUNT(Member_PCP_Spans.member_id) as CountMembers
FROM Member_PCP_Spans 
LEFT JOIN Providers ON Member_PCP_Spans.provider_id = Providers.id
LEFT JOIN Members ON Member_PCP_Spans.member_id = Members.id 
WHERE Member_PCP_Spans.end_date = '999999' AND
Providers.city != Members.city")` members seeing a provider practicing outside of the member's city. 

```{r Q9}
kable(dbGetQuery(DB, "SELECT COUNT(Member_PCP_Spans.member_id) as CountMembers
                            FROM Member_PCP_Spans 
                            LEFT JOIN Providers ON Member_PCP_Spans.provider_id = Providers.id
                            LEFT JOIN Members ON Member_PCP_Spans.member_id = Members.id 
                            WHERE Member_PCP_Spans.end_date = '999999' AND
                            Providers.city != Members.city")) 
```


## Part II. Based on the available information, which providers provide the best healthcare? What are limitations with the available data? What other information would you like to know? 

We are looking for the top five PCPs based on a metric that captures "the best customer service". Based on the available data, the pcp_rating variable from the Member_PCP_Spans table is the metric most aligned with our question. From the table below, we select the top five PCPs based on the highest mean of this metric and evaluating that observing that at all the selected providers had at least 30 ratings (range of 35 to 76). 

```{r Part2}
kable(dbGetQuery(DB, "WITH NewTable AS 
                 (SELECT provider_id, COUNT(pcp_rating) as NumObs, avg(pcp_rating) as MeanRating 
                 FROM Member_PCP_Spans WHERE pcp_rating IS NOT NULL 
                 GROUP BY provider_id ORDER BY MeanRating DESC LIMIT 5) 
                 SELECT provider_id, NumObs, MeanRating, name, city 
                 FROM NewTable 
                 INNER JOIN Providers ON Providers.id = NewTable.provider_id"))
```

Unfortunately, there are a number of missing values for several of the variables including the pcp_rating. In fact, `r dbGetQuery(DB, "SELECT COUNT(*) FROM Member_PCP_Spans WHERE pcp_rating IS NULL")` out of `r dbGetQuery(DB, "SELECT COUNT(*) FROM Member_PCP_Spans")` total entries in the table do not have a pcp_rating. 

A major limitation of this approach comes from the fact the PCP ratings are voluntary and given at a time when a member changes providers or plans. Therefore, there may be a selection bias meaning that the sample with pcp_rating's provided in the Member_PCP_Spans table are not representitative to the patient population as a whole. For example, there may be a bias in the sample towards patients that found a cheaper plan or changed plan due to unsatisfactory relationship with the PCP. You could imagine that an ideal provider would have many satsified patients who would not change provider or plan, which would not be captured by the pcp_variable. 

To make a more informed decision, additional data collection should be performed. Additional data could include a survey (preferably one that is mandatory or incentivized to minimize selection bias and be more population-based) given immediately (electronically or paper form) after a patient visit (to minize recall bias) that asks several questions preferably on a Likert scale including "How satisified are you with your current provider" (primary metric), "How likely are you to switch providers?", "Would you recommend your provider to a friend?", "Would you see your primary care provider again?". Similar questions are described in the book "The Innovator's Prescription: A Disruptive Solution for Health Care" by Clayton Christensen. 

Additionally, the provided data covers a year but there could be additional analyses if more longitudinal data is present. I would recommend considering secondary and more indirect measures as well. This could include identifying providers with the highest duration of patient-provider relationship or conversely, the providers with the smallest rate of member switching  after adjusting for patient risk score, average co-pay, length of employment, and other variable that might be causally related to the provider and outcome of interest. For example, a mixed-effects model (clustered by hospital or clinic) may be helpful to disentengle the effect of hospital or clinic and provider-patient satistification. 



